Loop Invariant Code Motion in PyPy's Tracing JIT

### 2011-06-16 talk/iwtc11/paper.tex
Use uppercase L for loops to prevent mixing up l and I.

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_0$($i_{0}$):
$i_1$ = $i_0$ + 1
print($i_1$)
jump($l_0$, $i_0$)
\end{lstlisting}

The first line is a label $l_0$ with argument $i_0$. Every label has a list of
arguments. The \lstinline{print} operation just prints its argument (it is not
an operation that PyPy's tracing JIT really supports, we just use it for this
example). The \lstinline{jump} operation jumps back to the beginning of the
trace, listing the new values of the arguments of the trace. In this case, the
new value of $i_0$ is $i_0$, making it a loop-invariant.

@@

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_0$($i_{0}$):
$i_1$ = $i_0$ + 1
print($i_1$)
jump($l_1$, $i_0$)

$l_1$($i_{0}$):
$i_2$ = $i_0$ + 1
print($i_2$)
jump($l_1$, $i_0$)
\end{lstlisting}

@@

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_0$($i_{0}$):
$i_1$ = $i_0$ + 1
print($i_1$)
jump($l_1$, $i_0$)

$l_1$($i_{0}$):
print($i_1$)
jump($l_1$, $i_0$)
\end{lstlisting}

This trace is malformed, because $i_1$ is used after the label $l_1$ without
being passed there, so we need to add $i_1$ as an argument to the label and pass
it along the \lstinline{jump}s:

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_0$($i_{0}$):
$i_1$ = $i_0$ + 1
print($i_1$)
jump($l_1$, $i_0$, $i_1$)

$l_1$($i_{0}$, $i_1$):
print($i_1$)
jump($l_1$, $i_0$, $i_1$)
\end{lstlisting}

@@

\begin{figure}
\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_0$($p_{0}$, $p_{1}$):
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = $i_{2} + i_{3}$
        $p_{5}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{5}$, intval, $i_{4}$)
jump($l_0$, $p_{0}$, $p_{5}$)
\end{lstlisting}
\caption{An Unoptimized Trace of the Example Interpreter}
\label{fig:unopt-trace}
\end{figure}

@@

The label of the loop is $l_0$ and is used by the jump instruction to
identify it's jump target.

@@

Loop peeling is achieved by appending an copy of the traced iteration at
the end of itselfe. See Figure~\ref{fig:overview}.
The first part (called \emph{preamble}) finishes with the jump the the second part
(called the \emph{peeled loop}). The second part end with the jump to itself. This way
the preamble will be executed only once while the peeled loop will
be used for every further iteration. New variable names have to be
introduced in the entire copied trace in order to maintian the SSA-property.
When applying the following optimizations to this two-iteration trace
some care has to taken as to how the arguments of the two
\lstinline{jump} operations and the input arguments of the peeled loop are
treated. It has to be ensured that the peeled loop stays a proper
trace in the sense that the operations within it only operates on
variables that are either among its input arguments 
or produced within the peeled loop. To ensure this we need
to introduce a bit of formalism.

@@

\begin{figure}
\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_0$($p_{0}$, $p_{1}$):
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = $i_{2}+i_{3}$
        $p_{5}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{5}$, intval, $i_{4}$)
jump($l_1$, $p_{0}$, $p_{5}$)

$l_1$($p_{0}$, $p_{5}$):
# inside f: y = y.add(step)
guard_class($p_{5}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{6}$ = get($p_{5}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{7}$ = get($p_{0}$, intval)
        $i_{8}$ = $i_{6}+i_{7}$
        $p_{9}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{9}$, intval, $i_{8}$)
jump($l_1$, $p_{0}$, $p_{9}$)
\end{lstlisting}
\caption{A peeled trace of the Example Interpreter}
\label{fig:peeled-trace}
\end{figure}

@@

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_1$($p_{0}$, $p_{5}$):
# inside f: y = y.add(step)
    # inside BoxedInteger.add
    $i_{6}$ = get($p_{5}$, intval)
        # inside BoxedInteger.add__int
        $i_{7}$ = get($p_{0}$, intval)
        $i_{8}$ = $i_{6}+i_{7}$
        $p_{9}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{9}$, intval, $i_{8}$)
jump($l_1$, $p_{0}$, $p_{9}$)
\end{lstlisting}

@@

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_0$($p_{0}$, $p_{1}$):
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = $i_{2}+i_{3}$
        $p_{5}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{5}$, intval, $i_{4}$)
jump($l_1$, $p_{0}$, $p_{5}$, $i_3$, $i_4$)

$l_1$($p_{0}$, $p_{5}$, $i_3$, $i_4$):
# inside f: y = y.add(step)
    # inside BoxedInteger.add
        # inside BoxedInteger.add__int
        $i_{8}$ = $i_{4}+i_{3}$
        $p_{9}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{9}$, intval, $i_{8}$)
jump($l_1$, $p_{0}$, $p_{9}$, $i_3$, $i_8$)
\end{lstlisting}

@@

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_0$($p_{0}$, $p_{1}$):
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = $i_{2}+i_{3}$
            # inside BoxedInteger.__init__
jump($l_1$, $p_{0}$, $i_{4}$)

$l_1$($p_{0}$, $i_{4}$):
# inside f: y = y.add(step)
    # inside BoxedInteger.add
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{7}$ = get($p_{0}$, intval)
        $i_{8}$ = $i_{4}+i_{7}$
            # inside BoxedInteger.__init__
jump($l_1$, $p_{0}$, $i_8$)
\end{lstlisting}

>>>>

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$L_0$($i_{0}$):
$i_1$ = $i_0$ + 1
print($i_1$)
jump($L_0$, $i_0$)
\end{lstlisting}

The first line is a label $L_0$ with argument $i_0$. Every label has a list of
arguments. The \lstinline{print} operation just prints its argument (it is not
an operation that PyPy's tracing JIT really supports, we just use it for this
example). The \lstinline{jump} operation jumps back to the beginning of the
trace, listing the new values of the arguments of the trace. In this case, the
new value of $i_0$ is $i_0$, making it a loop-invariant.

@@

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$L_0$($i_{0}$):
$i_1$ = $i_0$ + 1
print($i_1$)
jump($L_1$, $i_0$)

$L_1$($i_{0}$):
$i_2$ = $i_0$ + 1
print($i_2$)
jump($L_1$, $i_0$)
\end{lstlisting}

@@

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$L_0$($i_{0}$):
$i_1$ = $i_0$ + 1
print($i_1$)
jump($L_1$, $i_0$)

$L_1$($i_{0}$):
print($i_1$)
jump($L_1$, $i_0$)
\end{lstlisting}

This trace is malformed, because $i_1$ is used after the label $L_1$ without
being passed there, so we need to add $i_1$ as an argument to the label and pass
it along the \lstinline{jump}s:

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$L_0$($i_{0}$):
$i_1$ = $i_0$ + 1
print($i_1$)
jump($L_1$, $i_0$, $i_1$)

$L_1$($i_{0}$, $i_1$):
print($i_1$)
jump($L_1$, $i_0$, $i_1$)
\end{lstlisting}

@@

\begin{figure}
\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$L_0$($p_{0}$, $p_{1}$):
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = $i_{2} + i_{3}$
        $p_{5}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{5}$, intval, $i_{4}$)
jump($L_0$, $p_{0}$, $p_{5}$)
\end{lstlisting}
\caption{An Unoptimized Trace of the Example Interpreter}
\label{fig:unopt-trace}
\end{figure}

@@

The label of the loop is $L_0$ and is used by the jump instruction to
identify it's jump target.

@@

Loop peeling is achieved by appending an copy of the traced iteration at
the end of itselfe. See Figure~\ref{fig:overview} for an illustration.
The first part (called \emph{preamble}) finishes with the jump the the second part
(called the \emph{peeled loop}). The second part end with the jump to itself. This way
the preamble will be executed only once while the peeled loop will
be used for every further iteration. New variable names have to be
introduced in the entire copied trace in order to maintian the SSA-property.
When applying the following optimizations to this two-iteration trace
some care has to taken as to how the arguments of the two
\lstinline{jump} operations and the input arguments of the peeled loop are
treated. It has to be ensured that the peeled loop stays a proper
trace in the sense that the operations within it only operates on
variables that are either among its input arguments 
or produced within the peeled loop. To ensure this we need
to introduce a bit of formalism.

@@

\begin{figure}
\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$L_0$($p_{0}$, $p_{1}$):
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = $i_{2}+i_{3}$
        $p_{5}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{5}$, intval, $i_{4}$)
jump($L_1$, $p_{0}$, $p_{5}$)

$L_1$($p_{0}$, $p_{5}$):
# inside f: y = y.add(step)
guard_class($p_{5}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{6}$ = get($p_{5}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{7}$ = get($p_{0}$, intval)
        $i_{8}$ = $i_{6}+i_{7}$
        $p_{9}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{9}$, intval, $i_{8}$)
jump($L_1$, $p_{0}$, $p_{9}$)
\end{lstlisting}
\caption{A peeled trace of the Example Interpreter}
\label{fig:peeled-trace}
\end{figure}

@@

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$L_1$($p_{0}$, $p_{5}$):
# inside f: y = y.add(step)
    # inside BoxedInteger.add
    $i_{6}$ = get($p_{5}$, intval)
        # inside BoxedInteger.add__int
        $i_{7}$ = get($p_{0}$, intval)
        $i_{8}$ = $i_{6}+i_{7}$
        $p_{9}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{9}$, intval, $i_{8}$)
jump($L_1$, $p_{0}$, $p_{9}$)
\end{lstlisting}

@@

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$L_0$($p_{0}$, $p_{1}$):
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = $i_{2}+i_{3}$
        $p_{5}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{5}$, intval, $i_{4}$)
jump($L_1$, $p_{0}$, $p_{5}$, $i_3$, $i_4$)

$L_1$($p_{0}$, $p_{5}$, $i_3$, $i_4$):
# inside f: y = y.add(step)
    # inside BoxedInteger.add
        # inside BoxedInteger.add__int
        $i_{8}$ = $i_{4}+i_{3}$
        $p_{9}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{9}$, intval, $i_{8}$)
jump($L_1$, $p_{0}$, $p_{9}$, $i_3$, $i_8$)
\end{lstlisting}

@@

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$L_0$($p_{0}$, $p_{1}$):
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = $i_{2}+i_{3}$
            # inside BoxedInteger.__init__
jump($L_1$, $p_{0}$, $i_{4}$)

$L_1$($p_{0}$, $i_{4}$):
# inside f: y = y.add(step)
    # inside BoxedInteger.add
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{7}$ = get($p_{0}$, intval)
        $i_{8}$ = $i_{4}+i_{7}$
            # inside BoxedInteger.__init__
jump($L_1$, $p_{0}$, $i_8$)
\end{lstlisting}
